---
title: "BasalEstimates_W1W2W3"
author: "Sarah Donaldson"
date: "7/21/22"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
#LOAD PACKAGES AND SET DIRECTORIES

library(knitr)
library(psych)
library(dplyr)
library(here)
library(rio)
library(psych)
library(corrplot)
library(parameters)
library(nFactors)
library(ggplot2)
library(expss)
library(ltm)
library(GPArotation)
library(mice)
library(tidyverse)
library(janitor)

pacman::p_load("Amelia", "lme4" , "merTools", "ggplot2","tidyr","stringr","knitr","corrplot","data.table","readxl","gridExtra","dplyr", "psych","kableExtra","lavaan","jomo","Hmisc","DescTools", install = TRUE)

# cas_dir <- "/Volumes/psych-cog/dsnlab/TAG/"
# saliva_dir <- paste0(cas_dir,'behavior/Puberty/Saliva/cjm_cfm')
# puberty_dir <- paste0(cas_dir, 'behavior/Questionnaires/Puberty')
# options(digits=3)
```

###CHECK EFFECTS OF CONFOUNDS USING LMER (LONG DATASET WAVE 1)
```{r , include=FALSE}


Saliva <- import(here("data", "TAG_W1_Saliva_processed.csv"))
Saliva <- Saliva %>% mutate(week=as.numeric(week), SID=as.factor(SID))

#plot start_time distribution
Start_time <- Saliva %>% 
  dplyr::select(SID,week,start_time) %>% 
  arrange(start_time)

Start_time <- as.POSIXlt(Start_time$start_time)$hour

hist(Start_time, breaks=seq(0, 23), main="Start time (hour)")

#plot time_diff distribution
time_diff <- Saliva %>% 
  dplyr::select(SID,week,time_diff) %>% 
  arrange(time_diff)

hist(time_diff$time_diff,breaks=50, main="time between waking and starting (minutes)")

#Remove one negative time difference
Saliva$time_diff <- ifelse(Saliva$SID==042 & Saliva$week==1, NA, Saliva$time_diff)

#Test for time
TESTlmm_time <- Saliva %>% filter(complete.cases(sal_TEST_conc_ln_w,wake_up_time_num,time_diff,weekday))
TEST_null_time <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=TESTlmm_time, REML = F)
TEST_time <- lmer(sal_TEST_conc_ln_w ~ time_diff + (1 | SID), data=TESTlmm_time, REML = F)
TEST_time1 <- lmer(sal_TEST_conc_ln_w ~ wake_up_time_num + time_diff + (1 | SID), data=TESTlmm_time, REML = F)
TEST_time2 <- lmer(sal_TEST_conc_ln_w ~ wake_up_time_num + time_diff + weekday + (1 | SID), data=TESTlmm_time, REML = F)
anova(TEST_null_time,TEST_time,TEST_time1,TEST_time2)

#Test for meds
TESTlmm_med <- Saliva %>% filter(complete.cases(sal_TEST_conc_ln_w,medcode_1,medcode_2,medcode_3,medcode_5,
                                           medcode_6,medcode_7,medcode_13,medcode_16,medcode_17,medcode_18,
                                           medcode_19,medcode_20))
TEST_null_med <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=TESTlmm_med, REML = F)
TEST_medcode <- lmer(sal_TEST_conc_ln_w ~ medcode_1+medcode_2+medcode_3+medcode_5+medcode_6+medcode_7+medcode_13+medcode_16+medcode_17+medcode_18+medcode_19+medcode_20+ (1 | SID), data=TESTlmm_med, REML = F)
anova(TEST_null_med,TEST_medcode)

#Test for sick
TESTlmm_sick <- Saliva %>% filter(complete.cases(sal_TEST_conc_ln_w,sick_y_n))
TEST_null_sick <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=TESTlmm_sick, REML = F)
TEST_sick_y_n <- lmer(sal_TEST_conc_ln_w ~ sick_y_n + (1 | SID), data=TESTlmm_sick, REML = F)
anova(TEST_null_sick,TEST_sick_y_n)

```

##Wave 1 conclusions: time_diff (=time between waking and starting collection) matters, no other confounds.

###CHECK EFFECTS OF CONFOUNDS USING LMER (LONG DATASET WAVE 2)
```{r}

Saliva_withconfounds <- import(here("data", "TAG_W2_Saliva_processed.csv"))

Saliva_withconfounds <- Saliva_withconfounds %>% 
  mutate(week=as.numeric(week), SID=as.factor(SID))

#plot start_time distribution
W2Start_time <- Saliva_withconfounds %>% 
  dplyr::select(SID,week,start_time) %>% 
  arrange(start_time)

W2Start_time <- as.POSIXlt(W2Start_time$start_time)$hour

hist(W2Start_time, breaks=seq(0, 23), main="Start time (hour)")
#plot time_diff distribution
W2time_diff <- Saliva %>% select(SID,week,time_diff) %>% arrange(time_diff)
hist(W2time_diff$time_diff,breaks=50, main="time between waking and starting (minutes)")

#Test for time
W2TESTlmm_time <- Saliva_withconfounds %>% filter(complete.cases(sal_TEST_conc_ln_w,wake_up_time_num,time_diff,weekday))
W2TEST_null_time <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=W2TESTlmm_time, REML = F)
W2TEST_time <- lmer(sal_TEST_conc_ln_w ~ time_diff + (1 | SID), data=W2TESTlmm_time, REML = F)
W2TEST_time1 <- lmer(sal_TEST_conc_ln_w ~ wake_up_time_num + time_diff + (1 | SID), data=W2TESTlmm_time, REML = F)
W2TEST_time2 <- lmer(sal_TEST_conc_ln_w ~ wake_up_time_num + time_diff + weekday + (1 | SID), data=W2TESTlmm_time, REML = F)
anova(W2TEST_null_time,W2TEST_time,W2TEST_time1,W2TEST_time2)

#Test for meds
W2TESTlmm_med <- Saliva_withconfounds %>% filter(complete.cases(sal_TEST_conc_ln_w,medcode_1,medcode_4,medcode_5,medcode_6,                                           medcode_7,medcode_9,medcode_12,medcode_13,medcode_16,medcode_17,medcode_18,
                                           medcode_19,medcode_20))
W2TEST_null_med <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=W2TESTlmm_med, REML = F)
W2TEST_medcode <- lmer(sal_TEST_conc_ln_w ~ medcode_1+medcode_4+medcode_5+medcode_6+medcode_7+medcode_9+medcode_12+medcode_13+medcode_16+medcode_17+medcode_18+medcode_19+medcode_20+ (1 | SID), data=W2TESTlmm_med, REML = F)
anova(W2TEST_null_med,W2TEST_medcode)

#Test for sick
W2TESTlmm_sick <- Saliva_withconfounds %>% filter(complete.cases(sal_TEST_conc_ln_w,sick_y_n))
W2TEST_null_sick <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=W2TESTlmm_sick, REML = F)
W2TEST_sick_y_n <- lmer(sal_TEST_conc_ln_w ~ sick_y_n + (1 | SID), data=W2TESTlmm_sick, REML = F)
anova(W2TEST_null_sick,W2TEST_sick_y_n)

```

## Wave 2 conclusions: weekday matters and Medication matters 
#Med categories that matter: medcode_1 (glucocoticoid sprays/inhalers), medcode_4 (contraceptives), medcode_17 (antibiotics/antifungals)

###CHECK EFFECTS OF CONFOUNDS USING LMER (LONG DATASET WAVE 3)
```{r}

Saliva_withconfounds <- import(here("data", "TAG_W3_Saliva_processed.csv"))

Saliva_withconfounds <- Saliva_withconfounds %>% 
  mutate(week=as.numeric(week), SID=as.factor(SID))

#plot start_time distribution
W3Start_time <- Saliva_withconfounds %>% 
  dplyr::select(SID,week,start_time) %>% 
  arrange(start_time)

W3Start_time <- as.POSIXlt(W3Start_time$start_time)$hour

hist(W3Start_time, breaks=seq(0, 23), main="Start time (hour)")
#plot time_diff distribution
W3time_diff <- Saliva_withconfounds %>% 
  dplyr::select(SID,week,time_diff) %>% 
  arrange(time_diff)

hist(W3time_diff$time_diff,breaks=50, main="time between waking and starting (minutes)")

#Test for time
W3TESTlmm_time <- Saliva_withconfounds %>% filter(complete.cases(sal_TEST_conc_ln_w,wake_up_time_num,time_diff,weekday))
W3TEST_null_time <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=W3TESTlmm_time, REML = F)
W3TEST_time <- lmer(sal_TEST_conc_ln_w ~ time_diff + (1 | SID), data=W3TESTlmm_time, REML = F)
W3TEST_time1 <- lmer(sal_TEST_conc_ln_w ~ wake_up_time_num + time_diff + (1 | SID), data=W3TESTlmm_time, REML = F)
W3TEST_time2 <- lmer(sal_TEST_conc_ln_w ~ wake_up_time_num + time_diff + weekday + (1 | SID), data=W3TESTlmm_time, REML = F)
anova(W3TEST_null_time,W3TEST_time,W3TEST_time1,W3TEST_time2)

#Test for med
W3TESTlmm_med <- Saliva_withconfounds %>% filter(complete.cases(sal_TEST_conc_ln_w,medcode_12,medcode_13,medcode_17, medcode_18,medcode_19,medcode_20,medcode_3,medcode_4,medcode_5,medcode_6,medcode_7,medcode_9))
W3TEST_null_med <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=W3TESTlmm_med, REML = F)
W3TEST_medcode <- lmer(sal_TEST_conc_ln_w ~ medcode_3+medcode_4+medcode_5+medcode_6+medcode_7+medcode_9+medcode_12+medcode_13+medcode_17+medcode_18+medcode_19+medcode_20+ (1 | SID), data=W3TESTlmm_med, REML = F)
anova(W3TEST_null_med,W3TEST_medcode)

#Test for sick
W3TESTlmm_sick <- Saliva_withconfounds %>% filter(complete.cases(sal_TEST_conc_ln_w,sick_y_n))
W3TEST_null_sick <- lmer(sal_TEST_conc_ln_w ~ 1 + (1 | SID), data=W3TESTlmm_sick, REML = F)
W3TEST_sick_y_n <- lmer(sal_TEST_conc_ln_w ~ sick_y_n + (1 | SID), data=W3TESTlmm_sick, REML = F)
anova(W3TEST_null_sick,W3TEST_sick_y_n)

```
## Wave 3 conclusions: no significant confounds

##Decision: use all confounds that have a sign influence on at least one wave. This includes time_diff, weekday, medcode_1, medcode_4, medcode_17. Medcode_4 is not included in the imputation and basal estimate calculation for wave 1 below because it was 0 for every pp&sample.


###CONTROL FOR CONFOUNDS - WAVE 1
```{r}
#imputing using age, pds and confounds


#run imputation with Amelia
W1impDF <- Saliva %>% 
  dplyr::select(SID,
                week,
                weekday,
                wake_up_time_num,
                start_time_num,
                sick_y_n,
                medcode_1,
                medcode_17,
                ageS2,
                PUBcomp,
                sal_TEST_conc_ln_w)

aout_W1impDF <- amelia(W1impDF, idvars = c("SID",
                                           "sal_TEST_conc_ln_w"),  noms=c("weekday","sick_y_n","medcode_1","medcode_17"), ts="week", m = 50)

#create time diff after imputing wake_up_time and start_time
aout_W1impDF <- transform(aout_W1impDF, time_diff=start_time_num - wake_up_time_num)
#create list of imputed datasets - to be used for analyses
W1aout_list <- as.list(aout_W1impDF$imputations) 
#just to visually check the imputed dataset
W1aout <- as.data.frame(W1aout_list)

#now we run LMM with imputed datasets to extract random effects

#lmm of mulitply imputed datasets with merTools
W1TEST_lmm_imp <- lmerModList(sal_TEST_conc_ln_w ~ time_diff + weekday + sick_y_n + medcode_1 + medcode_17  + (1 | SID), data = W1aout_list)
modelFixedEff(W1TEST_lmm_imp)

#extract random effect and coefficients (which add fixed effect to the random effect) - this is calculated as a mean across the imputed datasets
W1TESTranef_imp <- lapply(W1TEST_lmm_imp, function(X) {
    ranef <- as.data.frame(ranef(X)) %>% select(grp,condval)
    coef <- as.data.frame(coef(X)$SID) %>% bind_cols(ranef)    
  })
W1TESTranef_imp <- rbindlist(W1TESTranef_imp) 
W1TESTranef_imp <- W1TESTranef_imp %>% group_by(grp) %>% 
  summarise(TESTcoef_imp = mean(`(Intercept)`),
            TESTranef_imp = mean(condval)) %>%
  rename(SID=grp) %>%
  arrange(SID)

```

###CONTROL FOR CONFOUNDS - WAVE 2
```{r}
#imputing using age, pds and confounds

#add age and PUBcomp to the dataframe
pubcomp <- read.csv(paste0(puberty_dir,'Allwaves_PubertyComposite.csv'))
pubcomp <- pubcomp  %>% filter(wave==2) %>% select(SID,age,PUBcomp) %>% mutate(SID=as.factor(SID))
Saliva_W2 <- Saliva_withconfounds %>% left_join(., pubcomp)

#run imputation with Amelia
W2impDF <- Saliva_W2 %>% 
  select(SID,week,weekday,wake_up_time_num,start_time_num,sick_y_n,medcode_1,medcode_4,medcode_17,age,PUBcomp,sal_DHEA_conc_ln_w,sal_TEST_conc_ln_w,sal_EST_conc_ln_w)

aout_W2impDF <- amelia(W2impDF, idvars = c("SID","sal_DHEA_conc_ln_w","sal_TEST_conc_ln_w","sal_EST_conc_ln_w"),  noms=c("weekday","sick_y_n","medcode_1","medcode_4","medcode_17"), ts="week", m = 50)

#create time diff after imputing wake_up_time and start_time
aout_W2impDF <- transform(aout_W2impDF, time_diff=start_time_num - wake_up_time_num)
#create list of imputed datasets - to be used for analyses
W2aout_list <- as.list(aout_W2impDF$imputations) 
#just to visually check the imputed dataset
W2aout <- as.data.frame(W2aout_list)

#now we run LMM with imputed datasets to extract random effects


#lmm of mulitply imputed datasets with merTools
W2TEST_lmm_imp <- lmerModList(sal_TEST_conc_ln_w ~ time_diff + weekday + sick_y_n + medcode_1 + medcode_4 + medcode_17  + (1 | SID), data = W2aout_list)
modelFixedEff(W2TEST_lmm_imp)

#extract random effect and coefficients (which add fixed effect to the random effect) - this is calculated as a mean across the imputed datasets
W2TESTranef_imp <- lapply(W2TEST_lmm_imp, function(X) {
    ranef <- as.data.frame(ranef(X)) %>% select(grp,condval)
    coef <- as.data.frame(coef(X)$SID) %>% bind_cols(ranef)    
  })
W2TESTranef_imp <- rbindlist(W2TESTranef_imp) 
W2TESTranef_imp <- W2TESTranef_imp %>% group_by(grp) %>% 
  summarise(TESTcoef_imp = mean(`(Intercept)`),
            TESTranef_imp = mean(condval)) %>%
  rename(SID=grp) %>%
  arrange(SID)

#calculate mean hormone levels before correction for confounds and combine with the basal estimates
Saliva_basal_W2 <- Saliva_W2 %>% select(SID, week, contains("conc_ln_w")) %>%
  group_by(SID) %>% summarise(TESTmean = mean(sal_TEST_conc_ln_w,na.rm=T)) %>%
 left_join(., W2TESTranef_imp)
```

###CONTROL FOR CONFOUNDS - WAVE 3
```{r}
#imputing using age, pds and confounds

#add age and PUBcomp to the dataframe
pubcomp <- import(here("data","Allwaves_PubertyComposite_updated.csv"))
pubcomp <- pubcomp  %>% filter(wave==3) %>% select(SID,age,PUBcomp) %>% mutate(SID=as.factor(SID))
Saliva_W3 <- Saliva_withconfounds %>% left_join(., pubcomp)

#run imputation with Amelia
W3impDF <- Saliva_W3 %>% 
  dplyr::select(SID,week,weekday,wake_up_time_num,start_time_num,medcode_4,medcode_17,age,PUBcomp,sal_TEST_conc_ln_w)

aout_W3impDF <- amelia(W3impDF, idvars = c("SID","sal_TEST_conc_ln_w"),  noms=c("weekday","medcode_4","medcode_17"), ts="week", m = 50)

#create time diff after imputing wake_up_time and start_time
aout_W3impDF <- transform(aout_W3impDF, time_diff=start_time_num - wake_up_time_num)
#create list of imputed datasets - to be used for analyses
W3aout_list <- as.list(aout_W3impDF$imputations) 
#just to visually check the imputed dataset
W3aout <- as.data.frame(W3aout_list)

#now we run LMM with imputed datasets to extract random effects


#lmm of mulitply imputed datasets with merTools
W3TEST_lmm_imp <- lmerModList(sal_TEST_conc_ln_w ~ time_diff + weekday + medcode_4 + medcode_17  + (1 | SID), data = W3aout_list)
modelFixedEff(W3TEST_lmm_imp)

#extract random effect and coefficients (which add fixed effect to the random effect) - this is calculated as a mean across the imputed datasets
W3TESTranef_imp <- lapply(W3TEST_lmm_imp, function(X) {
    ranef <- as.data.frame(ranef(X)) %>% select(grp,condval)
    coef <- as.data.frame(coef(X)$SID) %>% bind_cols(ranef)    
  })
W3TESTranef_imp <- rbindlist(W3TESTranef_imp) 
W3TESTranef_imp <- W3TESTranef_imp %>% group_by(grp) %>% 
  summarise(TESTcoef_imp = mean(`(Intercept)`),
            TESTranef_imp = mean(condval)) %>%
  rename(SID=grp) %>%
  arrange(SID)

#calculate mean hormone levels before correction for confounds and combine with the basal estimates
Saliva_basal_W3 <- Saliva_W3 %>% select(SID, week, contains("conc_ln_w")) %>%
  group_by(SID) %>% summarise(TESTmean = mean(sal_TEST_conc_ln_w,na.rm=T)) %>%
 left_join(., W3TESTranef_imp)
```

###CALCULATE PUBERTAL TIMING VARIABLES BY REGRESSING OUT AGE ALONG WITH THE CONFOUNDS
```{r}

# #### WAVE 1
# #run LMM with imputed datasets from above
# 
# #lmm of mulitply imputed datasets with merTools
# W1TESTtim_lmm_imp <- lmerModList(sal_TEST_conc_ln_w ~ ageS2 + time_diff + weekday + sick_y_n + medcode_1 + medcode_17  + (1 | SID), data = W1aout_list)
# modelFixedEff(W1TESTtim_lmm_imp)
# 
# #extract random effect and coefficients (which add fixed effect to the random effect) - this is calculated as a mean across the imputed datasets
# W1TESTrantim_imp <- lapply(W1TESTtim_lmm_imp, function(X) {
#     ranef <- as.data.frame(ranef(X)) %>% select(grp,condval)
#     coef <- as.data.frame(coef(X)$SID) %>% bind_cols(ranef)    
#   })
# W1TESTrantim_imp <- rbindlist(W1TESTrantim_imp) 
# W1TESTrantim_imp <- W1TESTrantim_imp %>% group_by(grp) %>% 
#   summarise(TESTcoef_imp = mean(`(Intercept)`),
#             TESTranef_imp = mean(condval)) %>%
#   rename(SID=grp) %>%
#   arrange(SID)
# 
# 
# #### WAVE 2
# #run LMM with imputed datasets from above
# 
# #lmm of mulitply imputed datasets with merTools
# W2TESTtim_lmm_imp <- lmerModList(sal_TEST_conc_ln_w ~ age + time_diff + weekday + sick_y_n + medcode_1 + medcode_17  + (1 | SID), data = W2aout_list)
# modelFixedEff(W2TESTtim_lmm_imp)
# 
# #extract random effect and coefficients (which add fixed effect to the random effect) - this is calculated as a mean across the imputed datasets
# W2TESTrantim_imp <- lapply(W2TESTtim_lmm_imp, function(X) {
#     ranef <- as.data.frame(ranef(X)) %>% select(grp,condval)
#     coef <- as.data.frame(coef(X)$SID) %>% bind_cols(ranef)    
#   })
# W2TESTrantim_imp <- rbindlist(W2TESTrantim_imp) 
# W2TESTrantim_imp <- W2TESTrantim_imp %>% group_by(grp) %>% 
#   summarise(TESTcoef_imp = mean(`(Intercept)`),
#             TESTranef_imp = mean(condval)) %>%
#   rename(SID=grp) %>%
#   arrange(SID)
# 
# #combine the hormone based timing estimates
# Saliva_hormone_timing_W1 <- W1TESTrantim_imp
# Saliva_hormone_timing_W2 <- W2TESTrantim_imp

```

```{r}
#combine wave 3 with others
W1W2_basal <- import(here("data","W1W2_Saliva_basalestimates.csv"))

Saliva_basal_W1W2W3 <- W1W2_basal %>% merge(., Saliva_basal_W3, by="SID", suffixes=c("_wave1","_wave2", "_wave3"),all=T)

# Saliva_hormone_timing_W1W2W3 <- Saliva_hormone_timing_W1W2 %>% merge(., Saliva_hormone_timing_W2W3, by="SID", suffixes=c("_wave1","_wave2"),all=T)

#save 
write.csv(Saliva_basal_W1W2W3, "/Users/sarahd/Dropbox/UOregon/Pfeiffer Lab/Dissssssss/dissertation_analyses/data/Saliva_basal_W1W2W3.csv")

```